A sheaf--function dictionary for algebraic tori over local fields

### 2013-10-03 shanghai.tex
Fixes, edits and restructuring of section 9

\begin{proof}
Let $\pi: B \to G^0$ be a discrete isogeny; set $A \ceq \ker \pi$.
  We will find a discrete isogeny $f: H\to G$
  such that that $H^0 = B$, $f^0 =\pi$ and
  $\pi_0(f) : \pi_0(H)\to \pi_0(G)$ is an isomorphism of component
  groups.  Namely, we will fit $\pi$ into the following diagram,
  \begin{equation}\label{extension-diagram}
  \begin{tikzcd}
  A \arrow[equal]{r} \dar & A \dar \\
  B \rar \dar[swap]{\pi} & H \rar \dar[swap]{f} & \pi_0(H) \arrow{d}[below,rotate=90]{\sim}[swap]{\pi_0(f)} \\
  G^0 \rar & G \rar & \pi_0(G),
  \end{tikzcd}
  \end{equation}
  where all rows and columns are exact and all maps are defined over
  $\Fq$.  We will do so by passing back and forth between group
  schemes over $\Fq$ and their $\bFq$-points.

  Set $\Gamma \ceq \Gal(\bFq/\Fq)$. Extensions of
  $G^0(\bFq)$ by $A(\bFq)$, such as $A(\bFq) \to B(\bFq) \to G^0(\bFq)$,
  correspond to classes in $\Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq))$.
  Similarly, extensions of $G(\bFq)$ by $A(\bFq)$ correspond to
  classes in $\Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq))$.  The map
  $G^0(\bFq) \to G(\bFq)$ induces the map
  \[
  \Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq))
  \]
  fitting into the long exact sequence derived from applying
  the functor $\Hom(\mbox{---}, A(\bFq))$ to $G^0(\bFq) \to G(\bFq) \to \pi_0(G)(\bFq)$:
  \[
  \Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq)) \to \Ext^2_{\ZZ[\Gamma]}(\pi_0(G)(\bFq), A(\bFq)).
  \]

  We now show that $\Ext^2_{\ZZ[\Gamma]}(\pi_0(G)(\bFq), A(\bFq)) = 0$.
%
  It suffices to show that $\pi_0(G)(\bFq)$ has a two-term resolution
  by injective $\ZZ[\Gamma]$-modules.  Since the action of $\Gamma$ on
  $\pi_0(G)(\bFq)$ is continuous, it factors through some finite quotient
  $\Gamma_m \ceq \Gamma / H_m$.  Any product of Dedekind domains has
  cohomological dimension $1$, \David{Add a citation for this fact} so $\pi_0(G)(\bFq)$ has a
  two term resolution by injective $\ZZ[\Gamma_m]$-modules.  Let $I$
  be one of these modules: we need to show that $I$ is injective as a
  $\ZZ[\Gamma]$-module.  Consider the universal diagram for injectivity:

  \[
  \begin{tikzcd}
  A \rar{a} \arrow[hook]{d} & I \\
  B
  \end{tikzcd}
  \]
  where $A$ and $B$ are arbitrary $\ZZ[\Gamma]$-modules.  Since $H_m$ acts trivially on $I$,
  $I_{H_m}A \ceq \langle (\gamma-1)a : \gamma \in H_m, a \in A \rangle \subseteq \ker(a).$
  Since $B / I_{H_m}B \cong B \otimes_A A / I_{H_m}A$, we get
  \[
  \begin{tikzcd}
  A \arrow[two heads]{r} \arrow[hook]{d} & A/I_{H_m}A \rar \arrow[hook]{d} & I \\
  B \arrow[two heads]{r} & B/I_{H_m}B \arrow[dashed]{ur} & {}
  \end{tikzcd}
  \]
  The dotted arrow exists since $I$ is injective as a
  $\ZZ[\Gamma_m]$-module and the outer diagram shows that $I$ is
  injective as a $\ZZ[\Gamma]$-module.

@@

@@

\begin{proof}
\Clifton{Turn this paragraph into a lemma. We use it below.}  Since $G$ is a smooth group scheme locally of finite type over $\Fq$, its
  identity component $G^0$ of $G$ is a connected smooth, 
  group scheme of finite type over $\Fq$, reduced over some finite extension of $\Fq$, by
   \cite{vdGeer-Moonen:AbelianVarieties}*{3.17}. 
	Since $\Fq$ is a finite field and hence perfect, $G^0$ is actually reduced over $\Fq$;
	see \cite{EGAIV2}*{Prop 6.4.1}.
	Since every group scheme over a field is separated 
	(see, for example, \cite{vdGeer-Moonen:AbelianVarieties}*{3.12}),  
	it follows that $G^0$ is a connected algebraic group.

    By Proposition~\ref{prop:connected}, every
  quasicharacter sheaf on $G^0$ is isomorphic to a 
  finite quasicharacter sheaf on $G^0$, 
  so to prove the proposition it suffices to show that every 
  finite quasicharacter sheaf on $G^0$ extends to a quasicharacter sheaf on $G$.

 Let $(\pi,\psi)$ be a finite quasicharacter sheaf on $G^0$.
 By Lemma~\ref{lemma:ext}, there is an extension of the 
 discrete isogeny $\pi : B \to G^0$ 
 (see Section~\ref{sec:finite} for the definition of a discrete isogeny) 
 to a discrete isogeny $f : H \to G^0$ 
 such that $\pi_0(f) : \pi_0(H)\to \pi_0(G)$ is an isomorphism.
 Then $(f,\psi)$ is a finite quasicharacter sheaf on $G$.
 Then $(f,\psi)\vert_{G^0} \iso (\pi,\psi)$,
 completing the proof that $\QC(G)\to \QC(G^0)$ is essentially surjective.
\end{proof}

>>>>

\begin{proof}
Let $\pi: B \to G^0$ be a discrete isogeny and set $A \ceq \ker \pi$.
  We will find a discrete isogeny $f: H\to G$
  such that that $H^0 = B$, $f^0 =\pi$ and
  $\pi_0(f) : \pi_0(H)\to \pi_0(G)$ is an isomorphism of component
  groups.  Namely, we will fit $\pi$ into the following diagram,
  \begin{equation}\label{extension-diagram}
  \begin{tikzcd}
  A \arrow[equal]{r} \dar & A \dar \\
  B \rar \dar[swap]{\pi} & H \rar \dar[swap]{f} & \pi_0(H) \arrow{d}[below,rotate=90]{\sim}[swap]{\pi_0(f)} \\
  G^0 \rar & G \rar & \pi_0(G),
  \end{tikzcd}
  \end{equation}
  where all rows and columns are exact and all maps are defined over
  $\Fq$.  We will do so by passing back and forth between group
  schemes over $\Fq$ and their $\bFq$-points.

  Set $\Gamma \ceq \Gal(\bFq/\Fq)$. Extensions of
  $G^0(\bFq)$ by $A(\bFq)$, such as $A(\bFq) \to B(\bFq) \to G^0(\bFq)$,
  correspond to classes in $\Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq))$.
  Similarly, extensions of $G(\bFq)$ by $A(\bFq)$ correspond to
  classes in $\Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq))$.  The map
  $G^0(\bFq) \to G(\bFq)$ induces the map
  \[
  \Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq))
  \]
  fitting into the long exact sequence derived from applying
  the functor $\Hom(\mbox{---}, A(\bFq))$ to $G^0(\bFq) \to G(\bFq) \to \pi_0(G)(\bFq)$:
  \[
  \Ext^1_{\ZZ[\Gamma]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Gamma]}(G^0(\bFq), A(\bFq)) \to \Ext^2_{\ZZ[\Gamma]}(\pi_0(G)(\bFq), A(\bFq)).
  \]
  Since finite fields have cohomological dimension $1$ \cite{serre:GaloisCohomology}*{\S II.3.3},
  $\Ext^2_{\ZZ[\Gamma]}(\pi_0(G)(\bFq), A(\bFq))$ vanishes \cite{cartan-eilenberg:HomologicalAlgebra}*{Thm. 2.6}.

@@

In order to use the results of Section~\ref{sec:connected}, we need to know that
identity component of $G$ is actually an algebraic group.

\begin{lemma} \label{lem:G0alg-grp}
If $G$ is a $\QC$-scheme then $G^0$ is a connected algebraic group.
\end{lemma}
\begin{proof}
 Since $G$ is a smooth group scheme locally of finite type over $\Fq$, its
 identity component $G^0$ of $G$ is a connected smooth, 
 group scheme of finite type over $\Fq$, reduced over some finite extension of $\Fq$
 \cite{vdGeer-Moonen:AbelianVarieties}*{3.17}. 
 Since $\Fq$ is a finite field and hence perfect, $G^0$ is actually reduced over $\Fq$
 \cite{EGAIV2}*{Prop 6.4.1}.  Since every group scheme over a field is separated 
 \cite{vdGeer-Moonen:AbelianVarieties}*{3.12},
 it follows that $G^0$ is a connected algebraic group.
\end{proof}

@@

\begin{proof}
  By Lemma~\ref{lem:G0alg-grp} and Proposition~\ref{prop:connected}, every
  quasicharacter sheaf on $G^0$ is isomorphic to a 
  finite quasicharacter sheaf on $G^0$, 
  so to prove the proposition it suffices to show that every 
  finite quasicharacter sheaf on $G^0$ extends to a quasicharacter sheaf on $G$.

 Let $(\pi,\psi)$ be a finite quasicharacter sheaf on $G^0$.
 By Lemma~\ref{lemma:ext}, there is an extension of the 
 discrete isogeny $\pi : B \to G^0$ to a discrete isogeny $f : H \to G^0$ 
 such that $\pi_0(f) : \pi_0(H)\to \pi_0(G)$ is an isomorphism.
 Then $(f,\psi)$ is a finite quasicharacter sheaf on $G$ and 
 $(f,\psi)\vert_{G^0} \iso (\pi,\psi)$.
\end{proof}
